Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-18 Paper/Paper.tex
added images and text

\institute{
	Anonymized for review \\
	\email{ananymous@insitution.edu}
	\and
	Anonymized for review \\
	\email{ananymous@insitution.edu}
}

@@

\begin{figure}[htbp]
	\centering
% 		\includegraphics[width=1.00\textwidth]{Images/modes.png}
	\caption{Capacitive proximity sensing modes \cite{Smith1996}}
	\label{fig:modes}
\end{figure}

@@

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in \cite{wimmer2006} who presented a toolkit for capacitive proximity sensing.

@@

\subsection{Capacitive Proximity Sensor}
power drain: 1mA when active
image : photo
image : board layout and schematics !

The capacitive proximity sensor for the wristband performs measurements in loading mode. The sensing circuit schematic is shown in Figure \ref{}. It is based on a timer that controls the charging and discharging cycles of the virtual capacitor that is built by the wristband electrode and the surrounding environment. The timer toggles from charge to discharge at the time when a threshold voltage at the capacitor is reached. This results in an astable operation with succeeding charge/discharge cycles. When the capacitance of the virtual capacitor increases, the charging time will also increase and vice-versa. Therefore, the capacitance is inversely proportional to the number of charging cycles in a given time span. In order to guard the sensor from measuring the capacitance to underlying objects, a shield is placed directly underneath the measuring electrode. The shield is driven with the same potential as the sensing electrode, such that the the capacitance between the two electrodes is negligible. Using this shielding method, the measured capacitance will only be slightly affected by the grounded underlying arm.

The proximity sensor board generates a clock signal with varying frequency depending on the charge and discharge cycles. The HedgeHog measures the resulting capacitance by counting the signal's edges over a gate time of approximately 9.5ms.

\begin{figure}[htbp]
	\centering
% 		\includegraphics[trim=1cm 1cm 1cm 1cm,clip,width=\textwidth]{Images/pseudocode.pdf}
	\caption{Measurement procedure}
	\label{fig:pseudocode}
\end{figure}
\section{Experiment}

@@

[EB: Hier plot von drinking und grasping into locker]

Figure \ref{fig:drinking} depicts the data that was recorded for the activity drinking water. The accelerometer data shows that there periodic up- and down-movements while the capacitive proximity sensor delivers data that is associated to the proximity of the table. Figure \ref{fig:opendoor} illustrates a grasp into a locker while preparing lunch. It is obvious that the locker's bottom surface that is located very closely to the wristband has an influence on the proximity sensor.

@@

\section*{Acknowledgments}
% We would like to thank the students of TU Darmstadt and the employees of Fraunhofer IGD that lent us their sitting and lying abilities for the evaluation. 
We would like to thank our evaluation participants from the TU Darmstadt and the Fraunhofer IGD.

>>>>

\institute{
	Anonymized for review \\
	\email{anonymous@insitution.edu}
	\and
	Anonymized for review \\
	\email{anonymous@insitution.edu}
}

@@

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{Images/modes.pdf}
	\caption{Three different capacitive proximity sensing modes can be distinguished. The sensing electrodes build up an electric field to objects in the environment, illustrated with a cloud. \cite{Smith1996}}
	\label{fig:modes}
\end{figure}

@@

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in \cite{Wimmer2007} who presented a toolkit for capacitive proximity sensing.

@@

\subsection{Capacitive Proximity Sensor}

The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode that aims to eliminate the influence of the arm that is a very proximate grounded object to the sensing electrode. The sensor draws a supply current of 1 mA at 3.3 V when active. This low power consumption qualifies the sensor for wearable proximity sensing applications.

image : photo

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{Images/schematic.pdf}
	\caption{The sensing circuit is based on a timer with an operational amplifier that acts as a voltage follower.}
	\label{fig:schematic}
\end{figure}

The sensing circuit schematic is shown in Figure \ref{fig:schematic}. It is based on a timer that controls the charging and discharging cycles of the virtual capacitor that is built by the sensing electrode and the surrounding environment. The timer toggles from charge to discharge at the time when a threshold voltage at the capacitor is reached. This results in an astable operation with succeeding charge/discharge cycles. When the capacitance of the virtual capacitor increases, the charging time will also increase and vice-versa. Therefore, the capacitance is inversely proportional to the number of charging cycles in a given time span. In order to guard the sensor from measuring the capacitance to underlying objects, a shield electrode is placed directly underneath the measuring electrode. The shield is driven with the same potential as the sensing electrode, such that the the capacitance between the two electrodes is negligible. Using this shielding method the measured capacitance will only be slightly affected by the grounded underlying arm.

\begin{figure}[htbp]
	\centering
 		\includegraphics[trim=1cm 1cm 1cm 1cm,clip,width=\textwidth]{Images/pseudocode.pdf}
	\caption{Measurement procedure}
	\label{fig:pseudocode}
\end{figure}

Figure \ref{fig:pseudocode} illustrates the operations required for a measurement cycle. The proximity sensor board generates a clock signal with varying frequency depending on the charge and discharge cycles. The HedgeHog measures the resulting capacitance by counting the signal's edges over a gate time of approximately 9.5ms. During that counting phase, the HedgeHog is sent to sleep in order to reduce power consumption. In the following, the HedgeHog applies run-length encoding on the measured data to reduce overhead and periodically logs the data to the integrated SD card.

\section{Experiment}

@@

Figure \ref{fig:drinking} depicts the data that was recorded for the activity drinking water. The accelerometer data shows that there periodic up- and down-movements while the capacitive proximity sensor delivers data that is associated to the proximity of the table. Figure \ref{fig:making-bread} illustrates the preperation of a bread with marmelade while preparing lunch. It is obvious that the capacitive proximity sensor's data provides valuable information about the distance of the wristband to the table.

@@

We have shown that current wearable inertial sensing devices for activity recognition can be significantly enhanced by integrating a capacitive proximity sensor into the wristband. Regarding the classication accuracy, we obtained an overall increase in correctly classified instances of TODO percent. Some classes, for instance ``preparing bread'' and ``drinking'', could benefit very much from proximity-related sensor data. For such classes the classication performance could be boosted by up to TODO percent.

The classification results could be improved by the usage of more than one sensing electrode in the wristband. For example the wristband could integrate four electrodes that are placed on each side of the arm. However, this will lead to smaller electrode surfaces that lead to a decrease of sensing distance. We think that these considerations are an important subject of future work. The sensor's power consumption can be decreased by shorter measurement windows and the choice of more energy efficient hardware components. Measuring pulse width lengths instead of counting the rising edges of the sensor's signal may reduce the required time needed for a measurement.

Capacitive proximity sensors represent a suitable new input modality for future activity recognition systems. The low power consumption as well as the invisible integration of a sensor into the wristband meets an essential requirement of wearable applications. Especially in AAL environments, these systems can help monitoring the course of chronic diseases by recognizing activities of daily life. This may improve the quality of life of persons affected and their caregivers.

\section*{Acknowledgments}
We would like to thank our evaluation participants from TU Darmstadt and Fraunhofer IGD.
